# Human and Economic loss analysis from Storm Data

## Synopsis

This is a report on the analysis of the various types of natural disasters in the US. This report finds which types of the calamity have causes maximum human fatalities and injuries. Analysis of economic damage by the various calamities has also been analyzed. Tornado, Excessive Heat, Flash Flood, Heat and Lightening causes the most human fatalities. Flood, Hurricane/Typhoon, Tornado, Storm SUrge and Hail are the top five type of disasters that have caused maximum economic damage. Economic damage has been calculated as sum of both property damage and crop damage.

```{r load_data}
data <- read.csv("repdata_data_StormData.csv.bz2")

head(data)
str(data)
```

## Data Processing

Quick check for the NA (missing) values. 
0 in output suggests that number of missing values in that particular column is 0.

```{r missing_value_check}
sum(is.na(data$PROPDMG))
sum(is.na(data$PROPDMGEXP))
sum(is.na(data$CROPDMG))
sum(is.na(data$CROPDMGEXP))
```

Below code is for processing the loss from three significant digits and exponential to actual decimal figures. We need to convert the data into actual numeric figures for summation.

```{r property_damage}
foo <- data$PROPDMGEXP == "K" | data$PROPDMGEXP == "k"
data$PROPDMG[foo] <- data$PROPDMG[foo] * 1000

foo <- data$PROPDMGEXP == "M" | data$PROPDMGEXP == "m"
data$PROPDMG[foo] <- data$PROPDMG[foo] * 1000000

foo <- data$PROPDMGEXP == "B" | data$PROPDMGEXP == "b"
data$PROPDMG[foo] <- data$PROPDMG[foo] * 1000000000

foo <- data$PROPDMGEXP == "H" | data$PROPDMGEXP == "h"
data$PROPDMG[foo] <- data$PROPDMG[foo] * 100

foo <- data$PROPDMGEXP == "+"
data$PROPDMG[foo] <- data$PROPDMG[foo] * 10

foo <- data$PROPDMGEXP >= "0" & data$PROPDMGEXP <= "8"
data$PROPDMG[foo] <- data$PROPDMG[foo] * 10
```

Same process for the crop loss

```{r crop_damage}
foo <- data$CROPDMGEXP == "K" | data$CROPDMGEXP == "k"
data$CROPDMG[foo] <- data$CROPDMG[foo] * 1000

foo <- data$CROPDMGEXP == "M" | data$CROPDMGEXP == "m"
data$CROPDMG[foo] <- data$CROPDMG[foo] * 1000000

foo <- data$CROPDMGEXP == "B" | data$CROPDMGEXP == "b"
data$CROPDMG[foo] <- data$CROPDMG[foo] * 1000000000

foo <- data$CROPDMGEXP == "H" | data$CROPDMGEXP == "h"
data$CROPDMG[foo] <- data$CROPDMG[foo] * 100

foo <- data$CROPDMGEXP == "+"
data$CROPDMG[foo] <- data$CROPDMG[foo] * 10

foo <- data$CROPDMGEXP >= "0" & data$CROPDMGEXP <= "8"
data$CROPDMG[foo] <- data$CROPDMG[foo] * 10
```

## Results

Quick check again for missing values

```{r missing_data_check_2}
sum(is.na(data$EVTYPE))
sum(is.na(data$FATALITIES))
sum(is.na(data$INJURIES))
```

Summing the data for various instances of each disaster

```{r health_effect}
fatalities <- aggregate(FATALITIES~EVTYPE,data,sum)
injuries <- aggregate(INJURIES~EVTYPE,data,sum)
health_effect <- merge(fatalities,injuries,by="EVTYPE")
```

TOP 20 fatality causing disasters

```{r health_result}
health_effect <- health_effect[order(health_effect$FATALITIES,health_effect$INJURIES,decreasing = TRUE),]
health_effect[1:20,]
```

Top 5 fatality causing disasters plot

```{r health_plot,fig.width=10}
with(health_effect[1:5,],barplot(FATALITIES,col = 1:5,names.arg=EVTYPE))
title("TOTAL FATALITIES IN US")
```

Calculating the economic cost of each disaster

```{r econ_damage}
econ_damage <- data$PROPDMG + data$CROPDMG
data <- cbind(data,econ_damage)
```

TOP 20 economic damage causing disasters

```{r econ_result}
economic_damage <- aggregate(econ_damage~EVTYPE,data,sum)
economic_damage <- economic_damage[order(economic_damage$econ_damage,decreasing = TRUE),]
economic_damage[1:20,]
```

TOP 5 economic damage causing disasters plot

```{r econ_plot,fig.width=10}
with(economic_damage[1:5,],barplot(econ_damage,col=1:5,names.arg = EVTYPE))
title("TOTAL ECONOMIC DAMAGE IN US")
```